Plotting NMDS plots in ggplot2
========================================================

The RMarkdown source to this file can be found [here](\Rmd\2014-04-13-vegan-ggplot2.Rmd)

One of my favorite packages in R is ggplot2, created by [Hadley Wickham](http://had.co.nz/).  This package allows you to create scientific quality figures of everything from shapefiles to NMDS plots.  I will run through a walkthrough in how to make a NMDS plot using the **vegan** package and **ggplot2**.  I am not going to go into the details into running the NMDS, as for this walkthrough I am making the assumption you already have a finalized output.  

```{r, results='hide', message=FALSE, warning=FALSE}
.libPaths("D:/RPackages")
```


### Load the libraries and get the data
```{r}
library(vegan)  #load the vegan package
library(ggplot2)  #load the ggplot2 package

data(varespec) #load the vegan package
head(varespec)  # look at the first 6 rows of the data
```
For this walkthrough I would like to assign a "group" to the each row of the data for illustration purposes.  Normally, your data will already belong to a grp and this next step will not be necessary.  

The basic process I will use to assign these groups is to to find the number of rows of the varespec data and then randomly sample half rows to group 'A' and the other half will be group 'B'.

```{r}
set.seed(123456)  #this will set the seed so that the random draw will be the same
nrow(varespec)

# create a grouping variable that has a length of 24, the same # of rows of varespec using the rep function
grp<-rep(NA,24)

#randomly sample 12 of those rows to belong in grp A
ind<-sample(1:nrow(varespec),12)

#assign those in ind to grp A
grp[ind]<-"A"
grp

#assign the NAs to grp B
grp[is.na(grp)]<-'B'

# Then take a look at the results
grp
```

### Run the NMDS using the **vegan** package
```{r}
vare.mds <- metaMDS(varespec)  #using all the defaults
vare.mds  #display the results

```


I am not a fan of using base R for graphics.  When you are in a pinch, they are ok to call but never hand in an assignment or attempt to submit
for a publication the default plots.

```{r fig.width=7, fig.height=6}
plot(vare.mds,type="t")
```

### Using **ggplot** for the NMDS plot

The first step is to extract the scores (the x and y coordinates of the site (rows) and species and add the grp variable we created before.  Once again the *grp* variable is not needed, I am just using it for illustration purposes.  For the data.scores, the result will be a 26 row x 4 column data.frame with the NMDS1 (x location) and NMDS2 (y location), designated by the site number and the group (grp).  The species.scores will be a 44 row by 3 column data.frame with the NMDS1 (x location), NMDS2 (y location), and species.  


```{r}

data.scores<-as.data.frame(scores(vare.mds))  #Using the scores function from vegan to extract the site scores and convert to a data.frame
data.scores$site<-rownames(data.scores)  # create a column of site names, from the rownames of data.scores
data.scores$grp<-grp #  add the grp variable created earlier
head(data.scores)  #look at the data

species.scores<-as.data.frame(scores(vare.mds,"species")) #Using the scores function from vegan to extract the species scores and convert to a data.frame
species.scores$species<-rownames(species.scores) # create a column of species, from the rownames of species.scores
head(species.scores) #look at the data

```
Now that we have the site and species scores, we can begin plotting with **ggplot2**.  First we will produce a plot like the base plot function.  

```{r fig.width=7, fig.height=6}
ggplot() + 
  geom_text(data=species.scores,aes(x=NMDS1,y=NMDS2,label=species),alpha=0.5) +  # add the species labels
  geom_point(data=data.scores,aes(x=NMDS1,y=NMDS2,shape=grp,colour=grp),size=3) + # add the point markers
  geom_text(data=data.scores,aes(x=NMDS1,y=NMDS2,label=site),size=6,vjust=0) +  # add the site labels
  scale_colour_manual(values=c("A" = "red", "B" = "blue")) +
  coord_equal() +
  theme_bw()

```

There are a couple of changes I like to make in the themes to make these a little nicer. 
```{r fig.width=7, fig.height=6}
ggplot() + 
  geom_text(data=species.scores,aes(x=NMDS1,y=NMDS2,label=species),alpha=0.5) +  # add the species labels
  geom_point(data=data.scores,aes(x=NMDS1,y=NMDS2,shape=grp,colour=grp),size=4) + # add the point markers
  geom_text(data=data.scores,aes(x=NMDS1,y=NMDS2,label=site),size=8,vjust=0,hjust=0) +  # add the site labels
  scale_colour_manual(values=c("A" = "red", "B" = "blue")) +
  coord_equal() +
  theme_bw() + 
  theme(axis.text.x = element_blank(),  # remove x-axis text
        axis.text.y = element_blank(), # remove y-axis text
        axis.ticks = element_blank(),  # remove axis ticks
        axis.title.x = element_text(size=18), # remove x-axis labels
        axis.title.y = element_text(size=18), # remove y-axis labels
        panel.background = element_blank(), 
        panel.grid.major = element_blank(),  #remove major-grid labels
        panel.grid.minor = element_blank(),  #remove minor-grid labels
        plot.background = element_blank())
```
Another way to look at these is to plot a hull around each of the groups.  To accomplish this, you can utilize the *chull* function.  In the below plot I dropped the site score labels.
```{r}

grp.a<-data.scores[data.scores$grp=="A",][chull(data.scores[data.scores$grp=="A",c("NMDS1","NMDS2")]),] # hull values for grp A
grp.b<-data.scores[data.scores$grp=="B",][chull(data.scores[data.scores$grp=="B",c("NMDS1","NMDS2")]),] # hull values for grp B

hull.data<-rbind(grp.a,grp.b)  #combine grp.a and grp.b
hull.data
```
and plot it out 
There are a couple of changes I like to make in the themes to make these a little nicer. 
```{r fig.width=7, fig.height=6}
ggplot() + 
  geom_polygon(data=hull.data,aes(x=NMDS1,y=NMDS2,fill=grp,group=grp),alpha=0.30) + # add the convex hulls
  geom_text(data=species.scores,aes(x=NMDS1,y=NMDS2,label=species),alpha=0.5) +  # add the species labels
  geom_point(data=data.scores,aes(x=NMDS1,y=NMDS2,shape=grp,colour=grp),size=4) + # add the point markers
  scale_colour_manual(values=c("A" = "red", "B" = "blue")) +
  coord_equal() +
  theme_bw() + 
  theme(axis.text.x = element_blank(),  # remove x-axis text
        axis.text.y = element_blank(), # remove y-axis text
        axis.ticks = element_blank(),  # remove axis ticks
        axis.title.x = element_text(size=18), # remove x-axis labels
        axis.title.y = element_text(size=18), # remove y-axis labels
        panel.background = element_blank(), 
        panel.grid.major = element_blank(),  #remove major-grid labels
        panel.grid.minor = element_blank(),  #remove minor-grid labels
        plot.background = element_blank())
```

**ggplot2** gives you a lot of flexibility in developing plots.  Whenever you are thinking of plotting with **ggplot2** you need to first get the data in a data.frame format.   Additionally, because **ggplot2** is based on the ["Grammar of Graphics"](http://books.google.com/books/about/The_Grammar_of_Graphics.html?id=_kRX4LoFfGQC) by Leland Wilkinson, you can only have two-axis.  Given that, each layer must have the same x and y colummn names.  In addition, the plots are built in layers.  If in the above plot, if you were to put the *geom_polygon* below the *geom_point* line then the hulls would cover up the points and text. 